Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-18 docs/manuscript/paper.tex
run a single example with dlogz=0.1

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[Kimpson]{Kimpson$^{1}$, Melatos, O'Leary, Evans, others, etc. etc. %
\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
%\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
\\
% List of institutions
$^{1}$Royal Astronomical Society, Burlington House, Piccadilly, London W1J 0BQ, UK}

@@

% Abstract of the paper
\begin{abstract}
This is an abstract
\end{abstract}

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the pulsar frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth. Indeed, the pulse TOAs are modulated harmonically at the GW frequency. We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section. The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
\subsubsection{Plane GW perturbation}
We consider a gravitational plane wave from a single,distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{[i(\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0)]} \ ,
\end{equation}
with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$. The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$. Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB). We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic. Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves over decadal timescales as \citep[e.g.][]{Zhu10},
\begin{equation}
	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_c$ is the chirp mass of the SMBHB, $f_{gw}(t=0)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ the length of the data timespan, which for PTAs is $\sim 10$ years. A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$. From Equation \eqref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52 \%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\subsubsection{Measurement equation}
In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
	\label{eq:momentum_shift}
\end{equation}
where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  neither $\boldsymbol{q}$ nor $d$ is not a function of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. This barycentering correction is typically applied when generating TOAs using e.g. {\sc tempo2} and related timing software. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression:
\begin{equation}
	f = p_{\alpha} u^{\alpha} \ . 
	\label{eq:freq_temporal}
\end{equation}
Due to the kinematical corrections from the barycentering process, 
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0) \ ,
\end{equation}
where the perturbations to $\boldsymbol{u}$ from the GW are of higher order and neglected. In this case the shift in the momentum from Equation \eqref{eq:momentum_shift} can be related to a shift in the frequency as,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t) \ ,
	\label{eq:measurement}
\end{equation}
where \textcolor{red}{TK: what is the best notation here to give $q^i$ an $(n)$ superscript?}
\begin{equation}
	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right) \ .
	\label{eq:g_func}
\end{equation}
It is will also prove instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
or in a trigonometric form 
\begin{align}
	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right) \right] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.

@@

	\begin{subfigure}[b]{1\textwidth}
		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
		\label{fig:25MB_bfs}
	\end{subfigure}
	\caption{Example application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$. We plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency. In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ and the state is estimated accurately. In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$ and the state is not estimated accurately. The top left panels show the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line). The middle left panels show the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta). The bottom left panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$. The right hand panels plot the distribution of the innovation over the total observation period. In (a) the greed/blue and red/magenta lines are completely overlaid.}
	\label{fig:four figures}
\end{figure*}

The Kalman filter \citep{Kalman1} is an algorithm for recovering the most likely evolution of a set of system state variables, $\boldsymbol{X}$, given some noisy measurements, $\boldsymbol{Y}$. It is a common technique in signal processing that has been applied successfully in neutron star astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}. In this work we use the linear Kalman filter which assumes a linear relation between $\boldsymbol{X}$ and $\boldsymbol{Y}$. Whilst the measurement equation, Equation \eqref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively. Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}. \newline

The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}. At each discrete timestep $i = 1, ... , M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \eqref{eq:measurement}. The measurement predictions can then be compared against the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation". The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$. The Gaussian log-likelihood can then be calculated at each timestep:
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
\end{eqnarray}
with the total log-likelihood simply the sum over all timesteps, i.e. 
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i \ . \label{eq:likelihood}
\end{eqnarray}
For fixed data $\boldsymbol{Y}$, $\mathcal{L}$ is a function of the static parameters of the model, i.e. $\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})$. Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are also functions of $\boldsymbol{\theta}$. If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the error in $\hat{\boldsymbol{X}}$, $\hat{\boldsymbol{Y}}$ is minimized and $\mathcal{L}$ is maximised. This is illustrated in Figure \ref{fig:6MB_BFS}; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity. The residuals correspond to random noise and are normally distributed. Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution. This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values. In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.

\subsection{Nested Sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}}
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work. Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is the ability to compute the evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches. Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We  follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. The primary tunable parameter in nested sampling is $n_{\rm live}$, where a greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}(1/\sqrt{n_{\rm live}})$. However the computational runtime scales as $\mathcal{O}(n_{\rm live})$ and so one must make a trade-off between uncertainty and runtime. \cite{Ashton2022} offer a rule-of-thumb where the minimum number of live points should be greater than the number of static parameters. The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$. We take $n_{\rm live} = 500$ for all results presented in this work.

@@

\subsection{Representative example}\label{sec:rep_example}
We initially characterise our method with a single example of a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source. The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected arbitrarily, subject to being astrophysically reasonable and representative. The injected $\boldsymbol{\theta}_{\rm gw}$ are summarised in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}. The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$ are as described in Section \ref{sec:synt_pta} and are also shown in Table \ref{tab:parameters_and_priors}.

@@

@@

>>>>

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[Kimpson]{Kimpson$^{1}$, Melatos, O'Leary, Evans, others, etc. etc.%
\thanks{Contact e-mail: \href{tom.kimpson@unimelb.edu.au}{tom.kimpson@unimelb.edu.au}}%
%\thanks{Present address: Science magazine, AAAS Science International, \mbox{82-88}~Hills Road, Cambridge CB2~1LQ, UK}%
\\
% List of institutions
$^{1}$Royal Astronomical Society, Burlington House, Piccadilly, London W1J 0BQ, UK}

@@

% Abstract of the paper
\begin{abstract}
	\textcolor{red}{TK: will obviously drop the bullet points, but useful for writing a concise abstract.}

	\begin{itemize}
		\item Continuous nHz gravitational waves from individual supermassive black hole binaries are a primary target for Pulsar Timing Array (PTA) observations.
		\item We present a new and complementary state-space approach to PTA data analysis.
		\item We show how to combine a Kalman filter and nested sampling to estimate te

	\end{itemize}

 using the combination of a Kalman filter and .

that is able to self consistenyl track the intrinsic timingnoise in PTA pulsars and disentangle it from the GW-induced TOA modulat

We illustrate how our method is able to recover the full posterior distributions of the N model parameters

We combine the to...

The method is tested...

bias observed....

\end{abstract}

@@

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the pulsar frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth. Indeed, the pulse TOAs are modulated harmonically at the GW frequency. We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section. The measurement equation is a key input into the Kalman filter in Section \eqref{sec:kalman_filter}
\subsubsection{Plane GW perturbation}
We consider a gravitational plane wave from a single,distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{ \{ i[\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0] \} } \ ,
\end{equation}
with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$. The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$. Throughout this paper we work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB). We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic. Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves over decadal timescales as \citep[e.g.][]{Zhu10},
\begin{equation}
	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_c$ is the chirp mass of the SMBHB, $f_{gw}(t=0)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ the length of the data timespan, which for PTAs is $\sim 10$ years. A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$. From Equation \eqref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52 \%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\subsubsection{Measurement equation}
In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
	\label{eq:momentum_shift}
\end{equation}
where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  neither $\boldsymbol{q}$ nor $d$ is a function of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. This barycentering correction is typically applied when generating TOAs using e.g. {\sc tempo2} \citep{tempo2} and related timing software. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\boldsymbol{u}$ is given by the coordinate-independent expression:
\begin{equation}
	f = p_{\alpha} u^{\alpha} \ . 
	\label{eq:freq_temporal}
\end{equation}
Due to the kinematical corrections from the barycentering process, 
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0) \ ,
\end{equation}
where the perturbations to $\boldsymbol{u}$ from the GW are of higher order and neglected. In this case the shift in the momentum from Equation \eqref{eq:momentum_shift} can be related to a shift in the frequency as,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t) \ ,
	\label{eq:measurement}
\end{equation}
where
\begin{equation}
	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i, \,(n)} q^{j, \,(n)}}{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)}}\right) \ .
	\label{eq:g_func}
\end{equation}
and $q^{i, \,(n)}$ labels the $i$-th coordinate component of the of the $n$-th pulsars position vector $\boldsymbol{q}^{(n)}$. It is also instructive to express Equation \eqref{eq:g_func} in a trigonometric form as,
\begin{align}
	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^{i, \,(n)} q^{j, \,(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
	& \times \Big\{ \cos\left[-\Omega t +\Phi_0\right] \nonumber \\
	&- \cos \left[-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right] \Big\} \ .
	\label{eq:g_func_trig}
\end{align}
Equations \eqref{eq:measurement} -- \eqref{eq:g_func_trig},  define a measurement equation that relates the intrinsic pulsar spin frequency to the frequency measured by an observer on Earth.

\textbf{MAYB ESUPRESS N ? rm \,}

\textbf{Introduce puslar terms dropping here insttea?}

@@

	\begin{subfigure}[b]{1\textwidth}
		\includegraphics[width=\textwidth]{images/Kalman_example_wrong_params}
		\caption{Kalman filter using incorrect estimate of the static parameters, $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$}
		\label{fig:25MB_bfs}
	\end{subfigure}
	\caption{Example application of a Kalman filter to estimate the intrinsic pulsar frequency state $f_{\rm p}(t)$ given a measured frequency $f_{\rm m}(t)$. We plot the ephemeris subtracted state $f'_{\rm p}(t) = f_{\rm p}(t) - f_{\rm em} (t)$ to better illustrate the stochastic wandering of the pulsar frequency. In (a) the Kalman filter is run using identically correct estimates of the static system parameters, $\hat{\boldsymbol{\theta}} = \boldsymbol{\theta}$ and the state is estimated accurately. In (b) the Kalman filter is run with $\hat{\boldsymbol{\theta}} \neq \boldsymbol{\theta}$ and the state is not estimated accurately. The top left panels show the true pulsar state $f'_{\rm p}(t)$ (blue line) and the state estimated by the Kalman filter, $\hat{f'}_{\rm p}(t)$ (green line). The middle left panels show the true $f_{\rm m}(t)$ (red line) and the measured frequency estimated  by the Kalman filter, $\hat{f}_{\rm m}(t)$ (magenta). The bottom left panels show the residual or the innovation $\epsilon(t) =f_{\rm m}(t) - \hat{f}_{\rm m}(t)$. The right hand panels plot the distribution of the innovation over the total observation period. In (a) the greed/blue and red/magenta lines are completely overlaid.  \textcolor{red}{TK: See Fig \ref{fig:kalmanexampleboth} in Appendix for an alternative version of this figure. Which do we prefer?}}
	\label{fig:four figures}
\end{figure*}

The Kalman filter \citep{Kalman1} is a Gaussian-Markov model used to algorithmically recover the most likely set of unknown system state variables, $\boldsymbol{X}$, given some noisy measurements, $\boldsymbol{Y}$. It is a common technique in signal processing that has been applied successfully in neutron star astrophysics \citep[e.g.][]{Meyers2021,Melatos2023}. In this work we use the linear Kalman filter which assumes a linear relation between $\boldsymbol{X}$ and $\boldsymbol{Y}$. Whilst the measurement equation, Equation \eqref{eq:measurement}, is non-linear in the static parameters it is linear in the state and measurement variables, $f_{\rm p}^{(n)}$ and $f_{\rm m}^{(n)}$ respectively. Extension to non-linear problems is straightforward using either an extended Kalman filter \citep{zarchan2000fundamentals}, the unscented Kalman filter \citep{882463van} or the particle filter \citep{Simon10}. \newline

The full set of Kalman recursions is presented in Appendix \ref{sec:kalman}. At each discrete timestep indexed by $ 1 \leq i  \leq M$, the Kalman filter returns an estimate of the state variables, $\hat{\boldsymbol{X}}_i$, and the covariance of those estimates, $\boldsymbol{P}_i$. The filter tracks the error in its predictions of $\boldsymbol{X}_i$ by projecting the state predictions back into measurement space, $\hat{\boldsymbol{X}}_i \mapsto \hat{\boldsymbol{Y}}_i$, via the measurement equation, Equation \eqref{eq:measurement}. The measurement predictions can then be compared against the true observed noisy measurements to get a residual $\boldsymbol{\epsilon}_i = \boldsymbol{Y}_i  - \hat{\boldsymbol{Y}}_i$, sometimes termed the ``innovation". The Kalman filter also calculates the uncertainty in $\boldsymbol{\epsilon}_i$ via the innovation covariance $\boldsymbol{S}_i = \langle \boldsymbol{\epsilon}_i \boldsymbol{\epsilon}_i^{\intercal} \rangle$. The Gaussian log-likelihood can then be calculated at each timestep:
\begin{eqnarray}
	\log \mathcal{L}_i =  -\frac{1}{2} \left (N \log 2 \pi + \log  \left | \boldsymbol{S}_i \right | + \boldsymbol{\epsilon}_i^{\intercal} \boldsymbol{S}_i^{-1}  \boldsymbol{\epsilon}_i \right ) \ ,
\end{eqnarray}
with the total log-likelihood simply the sum over all timesteps, i.e. 
\begin{eqnarray}
	\log \mathcal{L} =  \sum_{i=1}^{i=M} \log \mathcal{L}_i \ . \label{eq:likelihood}
\end{eqnarray}
For fixed data $\boldsymbol{Y}$, $\mathcal{L}$ is a function of the static parameters of the model, i.e. $\mathcal{L}$ = $\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})$. Similarly the estimates of the state and measurement variables, $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$, are also functions of $\boldsymbol{\theta}$. If the estimates of the system parameters, $\boldsymbol{\hat{\theta}}$ , that we pass to the Kalman filter are close to the true underlying parameters then the errors in $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{Y}}$ are	 minimized and $\mathcal{L}$ is maximised. This is illustrated in Figure \ref{fig:6MB_BFS}; given a timeseries of the measured pulsar frequency the Kalman filter is able to recover the evolution of the hidden state with high fidelity. The residuals correspond to random noise and are normally distributed. Conversely, if $\boldsymbol{\hat{\theta}}$ is not close to the true parameters then the filter is unable to recover the state evolution. This is demonstrated in Figure \ref{fig:25MB_bfs} where the Kalman filter was run with $\boldsymbol{\hat{\theta}}$ slightly perturbed away from the true values. In this case the filter cannot track the state variable accurately and the residuals are no longer Gaussian.

\subsection{Nested Sampling}\label{sec:nested_sampling}
We can use the likelihood returned by the Kalman filter, Equation \eqref{eq:likelihood}, in conjunction with likelihood-based inference methods to estimate the posterior distribution of $\boldsymbol{\theta}$ by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}}
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
In order to estimate the posterior distribution and the model evidence we use nested sampling \citep{Skilling} throughout this work. Nested sampling is an integration algorithm used for evaluating marginalised likelihood integrals, of the form given by Equation \eqref{eq:model_evidence}, that also returns samples from the posterior, $p(\boldsymbol{\theta} | \boldsymbol{Y})$. It does this by drawing a set of $n_{\rm live}$ live points from $\pi(\boldsymbol{\theta})$ and then iteratively replacing the live point with the lowest likelihood with a new live point drawn from $\pi(\boldsymbol{\theta})$, where the new live point is required to have a higher likelihood than the discarded point. The primary advantage of nested sampling is the ability to compute the evidence integral, which is key for model selection, and proves difficult without considerable extra cost for the usual Markov Chain Monte Carlo (MCMC) approaches. Nested sampling is also typically less computationally intensive than MCMC and can handle multi-modal problems \citep{Ashton2022}. For these reasons, it has enjoyed widespread adoption in the physical sciences, particularly within the cosmological community \citep{Mukherjee2006,Feroz2008,Handley2015}, but has also commonly been applied in astrophysics \citep{UltraNest2021}, particle physics \citep{proceedings2019033014} and materials science \citep{2009arXiv0906materials}. For a review of nested sampling we refer the reader to \cite{Buchner2021} and \cite{Ashton2022}. Multiple nested sampling algorithms and computational libraries exist. \citep[e.g.][]{Feroz2008,Feroz2009,Handley2015,dynesty2020,UltraNest2021}. For gravitational astrophysics it is common to use the \texttt{dynesty} sampler \citep{dynesty2020} via the \texttt{Bilby} \citep{bilby.507.2037A} front-end library. We  follow this precedent and use \texttt{Bilby} for all nested sampling Bayesian inference in this work. The primary tunable parameter in nested sampling is $n_{\rm live}$, where a greater number of live points is advantageous for large parameter spaces and multi-modal problems, whilst the uncertainties in the evidence and the posterior scale as $\mathcal{O}(1/\sqrt{n_{\rm live}})$. However the computational runtime scales as $\mathcal{O}(n_{\rm live})$ and so one must make a trade-off between uncertainty and runtime. \cite{Ashton2022} offer a rule-of-thumb where the minimum number of live points should be greater than the number of static parameters. The results presented in this work are generally robust to the choice of $n_{\rm live}$, subject to the requirement of $n_{\rm live} > 7 + 5N$. We take $n_{\rm live} = 500$ for all results presented in this work. \newline

@@

\subsection{Representative example}\label{sec:rep_example}
We initially characterise our method with a single example of a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source. The static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this injection are selected to be astrophysically reasonable and representative. The injected $\boldsymbol{\theta}_{\rm gw}$ are summarised in the ``Injected Values" column of Table \ref{tab:parameters_and_priors}. The static pulsar parameters $\boldsymbol{\theta}_{\rm psr}$ are as described in Section \ref{sec:synt_pta} and are also shown in Table \ref{tab:parameters_and_priors}.

@@

\textbf{CLARIFY re uninformative c.f loguniform}

\textbf{sin/cos priors realy uniform over cos(x)}

\textbf{referecne NANOgrav puslars and n index c.f. TAble ANTF}

\textbf{why did you choose NANOGrav you dick}

\textbf{priors on f, fdot, shrink to e.g. factor x ANTF error}

@@

\section{Alternative version of Fig 1}
